Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-08 paper/paper.tex
spelling

Next, we have the ``additive'' connectives. For these, we use the
non-linear types to define \emph{indexed} versions of the additive
disjunction $\oplus$ and additive conjunction $\&$ of linear
logic. These are like a linear version of the $\Sigma$ and $\Pi$
connectives of ordinary dependent type theory. The indexed $\&$ is
defined by a $\lambda$ and eliminated using projection. The indexed
$\oplus$ is like a \emph{weak} $\Sigma$ type: it has an injection
introduction rule $\sigma$, but its elimination rule is given by
\emph{pattern matching} rather than projections. Note that we can
define the more typical nullary and binary versions of these
connectives by using indexing over the empty and boolean type
respectively. We will freely use $\bot$ to refer to this empty
discunction and $\top$ to refer to the empty conjunction. Lastly, we
include a type $\equalizer {a}{f}{g}$ that we call the
\emph{equalizer} of $f$ and $g$. We think of this type as the
``subset'' of $A$ that satisfy the equation $f\,a\equiv g\,a$. If
these were non-linear types, such a type could be constructed using
the equality type as $\sum_{x:X} f\,x=_Y g\,x$, but this construction
can't be used because it uses a \emph{dependent} equality type. We
note that the equalizer type is not used in defining any of our
parsers or formal grammars, but instead will be used shortly to derive
an induction principle for proving equalities of functions whose
domain is an inductive type.

@@

Next, we introduce one of the most complex and important linear
type constructors of our development, \emph{indexed inductive linear
types} $\mu F$. To encode these, we first add a non-linear type
$\SPF\,X$ of \emph{strictly-positive functorial} linear type
expressions indexed by a non-linear type $X$. We give the rules for
strictly-positive functors in \Cref{fig:spf}. First, we add operations
for a recursive reference $\Var{x}$ the constant functor, $K$, tensor
products and additive conjunction and disjunction. As an example,
\max{TODO: make an example}.  Next we add an operation $\el(F)$ which
interprets our syntactic codes as an actual type constructor, and
define it for each of the operations given above. Lastly, we give a
\emph{functorial action} $\map(F)$ for each of the codes.\max{TODO:
  find a citation for this kind of approach. Container papers work
  this way for instance. I think there is a linear container paper
  that uses this appraoch}

@@

\section{Formal Grammar Theory in \theoryname}
\label{sec:applications}
This section explores the applications of \theoryabbv to conducting formal
gramamr theory. We demonstrate that several classical notions and constructions
integral to the theory of formal languages are faithfully represented
in \theoryname. By encoding well-established formal grammar concepts, we ensure
that our framework remains grounded in the
foundational principles of formal language theory while opening the door to
compositional formal verification of parsers.

@@

\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$ is
unabmiguous if there is at most one way to transform parses of any other grammar
$\B$ into parse of $\A$.

Note that this definition of unambiguity is preserved by strong equivalence, so a common proof technique for showing $\A$ is unambiguous is by proving $\A$
is strongly equivalent to an unambguous grammar $\B$. In this style of proof, we
build the parse transformers $\f : \ltonl{(\A \lto \B)}$ and
$\g : \ltonl{(\B \lto \A)}$, then we have the obligation to show
$\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$. However, because $\B$ is
unambiguous there is only one parse transformer $\ltonl{(\B \lto \B)}$, so
$\f \circ \g : \ltonl{(\B \lto \B)}$ and $\ident : \ltonl{(\B \lto \B)}$ are equal.
Thus, in this type of proof, we really only have the obligation of building each of $\f$
and $\g$ and demonstrating that $\g \circ \f \equiv \ident$.

@@

\paragraph{Nondeterministic Finite Automata} In \cref{sec:type-theory-examples}, we saw
one particular NFA corresponding to the regular expression
$(a^{*} \otimes b) \oplus c$. The definition of the linear type of traces
through this
NFA, as
presented in \cref{fig:nfainductive}, inspects the concrete structure of the
NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
luxury of inspecting a concrete defintion. Therefore, we abstract over the
specification of an arbitrary NFA and define traces with respect to that
specification.

@@

\steven{TODO make the agda logo clickable, and mention its meaning somewhere}
\newcommand{\Agda}{\agdalogo}
\begin{theorem}[Determinization, \Agda]
  Given an NFA $N$, there exists a DFA $D$ such that
  $\Trace_{N}~\N.\init~\true$ is weakly equivalent to $\Trace_{D}~\D.\init~\true$.
\end{theorem}
\newcommand{\X}{\texttt{X}}
\newcommand{\Y}{\texttt{Y}}
\newcommand{\x}{\texttt{x}}
\newcommand{\y}{\texttt{y}}
\begin{proof}
  We present a variant of Rabin and Scotts's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
  the DFA $\D$. Define the
  states of $\D$ to be the $\mathbb{P}_{\varepsilon}(\N.\states)$ --- the type
  of $\epsilon$-closed\footnote{A subset of states $\X$ is $\epsilon$-closed if
    for every $\s \in \X$ and $\epsilon$-transition $\s \overset{\epsilon}{\to} \s'$ we have $\s' \in \X$.} subsets of
  $\N.\states$. A subset is accepting in $\D$ if it contains an accepting state
  from $\N$. Construct the initial state of $\D$ as the
  $\epsilon$-closure of $\N.\init$. Lastly define the transtion function
  of $\D$ to send the subset $X$ under the character $\c$ to the
  $\epsilon$-closure of all the states reachable from $X$ via a transition
  labelled with the character $\c$.

@@

If the NFA trace is defined via $\cons$ for a labeled transition
$\texttt{trans} : \N.\transitions$ such that $N.\src~\texttt{trans} \in \X$, then
we first decompose the trace into a parse of $\N.\labelt~\texttt{trans}$ and a
trace starting from $\N.\dst~\texttt{trans}$. We then recursively call $\NtoD$
on this trace which returns a family of traces parametrized by $\epsilon$-closed
subsets containing $N.\dst~\texttt{trans}$. We instantiate this family at the
set of states
$\D.\delta~(\N.\labelt~\texttt{trans})~X$. Because $\N.\dst~\texttt{trans}$ is
reachable from $\N.\src~\texttt{trans}$ by a single transition labeled by
$\N.\labelt~\texttt{trans}$ and $\N.\src~\texttt{trans} \in X$ we have that
$\N.\dst~\texttt{trans} \in \D.\delta~(\N.\labelt~\texttt{trans})~X$. Therefore
we can linearly combine the parse of the label and the trace built from the
recursive call to get a trace in $\D$ rooted at $X$.

Lastly, if the NFA trace is defined via $\epsilon\cons$ for an $\epsilon$-transition
$\epsilon\texttt{trans} : \N.\epstransitions$ such that
$\N.\epssrc~\epsilon\texttt{trans} \in \X$, then the construction is quite
similar. In this case we again recursivley call $\NtoD$, except afterwards we
instantiate the family of grammars at the set $X$ itself. We may do this because
all states in $\D$ are taken to be $\epsilon$-closed, therefore
$\N.\epsdst~\epsilon\texttt{trans} \in \X$ because
$\N.\epssrc~\epsilon\texttt{trans}$ is in $X$. This concludes the construction of $\NtoD$.

@@

Next if the trace in $\D$ is defined via transition
$\X \overset{\c}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
$\s \in \D.\delta~\c~\X$. Recursively, $\DtoN$ sends the trace rooted at
$\D.\delta~\c~\X$ to one rooted at $\s$. By definition of $\D.\delta~\c~\X$,
there exists some $\s' \in \X$ such that $\s'$ tranitions to $\s$ by some number of
$\epsilon$-transitions followed by a transition labeled by $\c$.
We order all
such traces of this form and choose the smallest one, which is then linearly
combined with the parse of $\c$ to build an $\N$ trace. This concludes the
definition of $\DtoN$, thus showing that $\Trace_{\N}~\N.\init~\true$ and $\Trace_{\D}~\D.\init~\true$ are
weakly equivalent.
\end{proof}

@@

$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$
and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
the empty string is balanced, and the $\balanced$ constrcutor builds a balanced
parse by wrapping an already balanced parse in an additional set of parentheses
then following it with another balanced parse.

@@

\subsection{Turing Machines}
\newcommand{\Turing}{\mathtt{Turing}}
\newcommand{\ttt}{\mathtt{t}}
\newcommand{\T}{\mathtt{T}}
\newcommand{\acc}{\mathtt{acc}}
\newcommand{\rej}{\mathtt{rej}}
\newcommand{\Tape}{\mathtt{Tape}}
\newcommand{\parseTape}{\mathtt{parseTape}}
\newcommand{\snoc}{\mathtt{snoc}}
\steven{This subsection is still a WIP}
%% 5. Expressiveness: Turing Machines
The expressive power of \theoryabbv spans the entire Chomsky hierachy, as we may
encode the acccepting traces of Turing machines as linear types.

@@

where $\mathsf{TuringTrace}$ is a nonlinear type denoting terminating traces
through the Turing machine. Note that $\mathsf{run}$ returns a $\mathsf{Maybe}$
type in a fuel parameter. Turing machines may not terminate. If they do
terminate within $\mathsf{fuel}$-many steps, then we return the trace along with
the current tape and a boolean denoting if the trace is accepting or rejcting.
If the trace does not terminate within the alloted amount of fuel, then we
timeout and return nothing.

@@

Parse transformers have an obvious notion of composition and identity
transformer. This defines a \emph{category}.
\begin{definition}
  Define $\Grammar$ to be the category whose objects are formal
  grammars and morphisms are parse transformers.
\end{definition}
This category is equivalent to the slice category $\Set/\Sigma^*$ and
as such is very well-behaved. It is complete, co-complete, cartesian
closed and carries a monoidal biclosed structure. We will use these
structures to model the linear types, terms and equalities in
\theoryabbv.
%% \begin{theorem}
%%   The category $\Grammar$ is complete, co-complete and cartesian
%%   closed. Furthermore, it is monoidal biclosed with the monoidal
%%   product given by
%%   \[ (A \otimes B) w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A w_1 \wedge b \in B w_2\} \]
%%   and unit by
%%   \[ I w = \{ () \pipe w = \varepsilon \} \]
%% \end{theorem}

@@

For reasons of brevity, we present only a selection of cases in
\Cref{fig:semantics}. The full details are in the appendix.\max{todo:
  put the full details in the appendix}.
\max{TODO: how to describe $\mu$ succinctly?}\max{make sure $\lto$ and $\tol$ are right}
\begin{figure}
  \begin{align*}
    \sem{c}\gamma\,w &= \{ c | w = c\}\\
    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
    \sem{\SPF\,X}\gamma &= \{ F : \textrm{Functor}(\textrm{Grammar}^{\sem{X}\gamma}, \textrm{Grammar}) \pipe F \textrm{ preserves $\omega$-colimits.} \}\\
    \sem{\el(F)}\gamma\,G &= \sem{A}\gamma\,G\\
    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:adamek}}
  \end{align*}
  \caption{Grammar Semantics}
\end{figure}

@@

\max{todo: spelling}
\begin{theorem}[Adamek]
  If a functor $F : \mathcal C \to \mathcal C$ preserves
  $\omega$-colimits and $\mathcal C$ has an initial object and
  $\omega$-colimits, then $F$ has an initial algebra given by the
  colimit of the chain
  \[ 0 \to F(0) \to F(F(0)) \to \cdots \]
\end{theorem}

@@

\steven{Should discussion of this connection to lambek calculus be spread
  throughout the paper?}

Lambek later explicitly connects his syntactic calculus to the structure of a
non-commutative biclosed monoidal category \cite{lambek1988categorial}. There is
a rich line of subsequent research after these revelations by Lambek, including
the study of pregroup grammars
\cite{coeckeMathematicalFoundationsCompositional2010}, the presentation of
abstract categorial grammars \cite{degrooteAbstractCategorialParsing2015}, and
even Lambek calculus with dependent types \cite{luoLambekCalculusDependent}. As
far as the authors can tell, these works seem to be primarily interested in natural
language semantics in the style of Montague grammars, rather than parsing formal
languages. That is, they seem to try to give a functorial semantic
interpretation of natrual language.

@@

Also mention here that \cite{luoLambekCalculusDependent} seems similar at first
glance, but isn't as similar as it might look because the judgements and
dependent types seem to
take on a different interpretation of ours. Say something here like, ``as we
understand it, it differs in x,y,z ways''

\steven{TODO Edit the above}
\steven{Even though the above sounds bad, are there lambek/categorial things that are missed?}

@@

\steven{TODO add lambek calculus/categorial grammars here}

@@

    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
% %
%     \and
%
%     \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
% %
%     \and
% %
%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} {e \otimes e'} e'' \equiv e'' \{ e/a , e'/b \} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} e c : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(M, e)} {e'} \equiv e' \{ M/x , e/a \} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
    \and
    %
    \inferrule
    {~}
    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerin {e}} \equiv e : A}
    %
    \and
    %
    \inferrule
    {~}
    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e : A}
\end{mathpar}
  \caption{Judgmental equality for linear terms}
  \label{fig:jdgeq}
\end{figure}
\steven{Do we need inverse for $\ltonl A$ in judgemental equality figure?}
\steven{Do we need to say anything about the equalizer proofs being unique in
  judgemental equality figure?}

>>>>

Next, we have the ``additive'' connectives. For these, we use the
non-linear types to define \emph{indexed} versions of the additive
disjunction $\oplus$ and additive conjunction $\&$ of linear
logic. These are like a linear version of the $\Sigma$ and $\Pi$
connectives of ordinary dependent type theory. The indexed $\&$ is
defined by a $\lambda$ and eliminated using projection. The indexed
$\oplus$ is like a \emph{weak} $\Sigma$ type: it has an injection
introduction rule $\sigma$, but its elimination rule is given by
\emph{pattern matching} rather than projections. Note that we can
define the more typical nullary and binary versions of these
connectives by using indexing over the empty and boolean type
respectively. We will freely use $\bot$ to refer to this empty
disjunction and $\top$ to refer to the empty conjunction. Lastly, we
include a type $\equalizer {a}{f}{g}$ that we call the
\emph{equalizer} of $f$ and $g$. We think of this type as the
``subset'' of $A$ that satisfy the equation $f\,a\equiv g\,a$. If
these were non-linear types, such a type could be constructed using
the equality type as $\sum_{x:X} f\,x=_Y g\,x$, but this construction
can't be used because it uses a \emph{dependent} equality type. We
note that the equalizer type is not used in defining any of our
parsers or formal grammars, but instead will be used shortly to derive
an induction principle for proving equalities of functions whose
domain is an inductive type.

@@

Next, we introduce one of the most complex and important linear
type constructors of our development, \emph{indexed inductive linear
types} $\mu F$. To encode these, we first add a non-linear type
$\SPF\,X$ of \emph{strictly-positive functorial} linear type
expressions indexed by a non-linear type $X$. We give the rules for
strictly-positive functors in \Cref{fig:spf}. First, we add operations
for a recursive reference $\Var{x}$ the constant functor, $K$, tensor
products and additive conjunction and disjunction. As an example,
\max{TODO: make an example}.  Next we add an operation $\el(F)$ which
interprets our syntactic codes as an actual type constructor, and
define it for each of the operations given above. Lastly, we give a
\emph{functorial action} $\map(F)$ for each of the codes.\max{TODO:
  find a citation for this kind of approach. Container papers work
  this way for instance. I think there is a linear container paper
  that uses this approach}

@@

\section{Formal Grammar Theory in \theoryname}
\label{sec:applications}
This section explores the applications of \theoryabbv to conducting formal
grammar theory. We demonstrate that several classical notions and constructions
integral to the theory of formal languages are faithfully represented
in \theoryname. By encoding well-established formal grammar concepts, we ensure
that our framework remains grounded in the
foundational principles of formal language theory while opening the door to
compositional formal verification of parsers.

@@

\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$ is
unambiguous if there is at most one way to transform parses of any other grammar
$\B$ into parse of $\A$.

Note that this definition of unambiguity is preserved by strong equivalence, so a common proof technique for showing $\A$ is unambiguous is by proving $\A$
is strongly equivalent to an unambiguous grammar $\B$. In this style of proof, we
build the parse transformers $\f : \ltonl{(\A \lto \B)}$ and
$\g : \ltonl{(\B \lto \A)}$, then we have the obligation to show
$\f \circ \g \equiv \ident$ and $\g \circ \f \equiv \ident$. However, because $\B$ is
unambiguous there is only one parse transformer $\ltonl{(\B \lto \B)}$, so
$\f \circ \g : \ltonl{(\B \lto \B)}$ and $\ident : \ltonl{(\B \lto \B)}$ are equal.
Thus, in this type of proof, we really only have the obligation of building each of $\f$
and $\g$ and demonstrating that $\g \circ \f \equiv \ident$.

@@

\paragraph{Nondeterministic Finite Automata} In \cref{sec:type-theory-examples}, we saw
one particular NFA corresponding to the regular expression
$(a^{*} \otimes b) \oplus c$. The definition of the linear type of traces
through this
NFA, as
presented in \cref{fig:nfainductive}, inspects the concrete structure of the
NFA.\ When generalizing from this example to an arbitrary NFA, we do not have the
luxury of inspecting a concrete definition. Therefore, we abstract over the
specification of an arbitrary NFA and define traces with respect to that
specification.

@@

\steven{TODO make the Agda logo clickable, and mention its meaning somewhere}
\newcommand{\Agda}{\agdalogo}
\begin{theorem}[Determinization, \Agda]
  Given an NFA $N$, there exists a DFA $D$ such that
  $\Trace_{N}~\N.\init~\true$ is weakly equivalent to $\Trace_{D}~\D.\init~\true$.
\end{theorem}
\newcommand{\X}{\texttt{X}}
\newcommand{\Y}{\texttt{Y}}
\newcommand{\x}{\texttt{x}}
\newcommand{\y}{\texttt{y}}
\begin{proof}
  We present a variant of Rabin and Scott's classic powerset construction \cite{rabinFiniteAutomataTheir1959}. First, we build
  the DFA $\D$. Define the
  states of $\D$ to be the $\mathbb{P}_{\varepsilon}(\N.\states)$ --- the type
  of $\epsilon$-closed\footnote{A subset of states $\X$ is $\epsilon$-closed if
    for every $\s \in \X$ and $\epsilon$-transition $\s \overset{\epsilon}{\to} \s'$ we have $\s' \in \X$.} subsets of
  $\N.\states$. A subset is accepting in $\D$ if it contains an accepting state
  from $\N$. Construct the initial state of $\D$ as the
  $\epsilon$-closure of $\N.\init$. Lastly define the transition function
  of $\D$ to send the subset $X$ under the character $\c$ to the
  $\epsilon$-closure of all the states reachable from $X$ via a transition
  labelled with the character $\c$.

@@

If the NFA trace is defined via $\cons$ for a labeled transition
$\texttt{trans} : \N.\transitions$ such that $N.\src~\texttt{trans} \in \X$, then
we first decompose the trace into a parse of $\N.\labelt~\texttt{trans}$ and a
trace starting from $\N.\dst~\texttt{trans}$. We then recursively call $\NtoD$
on this trace which returns a family of traces parameterized by $\epsilon$-closed
subsets containing $N.\dst~\texttt{trans}$. We instantiate this family at the
set of states
$\D.\delta~(\N.\labelt~\texttt{trans})~X$. Because $\N.\dst~\texttt{trans}$ is
reachable from $\N.\src~\texttt{trans}$ by a single transition labeled by
$\N.\labelt~\texttt{trans}$ and $\N.\src~\texttt{trans} \in X$ we have that
$\N.\dst~\texttt{trans} \in \D.\delta~(\N.\labelt~\texttt{trans})~X$. Therefore
we can linearly combine the parse of the label and the trace built from the
recursive call to get a trace in $\D$ rooted at $X$.

Lastly, if the NFA trace is defined via $\epsilon\cons$ for an $\epsilon$-transition
$\epsilon\texttt{trans} : \N.\epstransitions$ such that
$\N.\epssrc~\epsilon\texttt{trans} \in \X$, then the construction is quite
similar. In this case we again recursively call $\NtoD$, except afterwards we
instantiate the family of grammars at the set $X$ itself. We may do this because
all states in $\D$ are taken to be $\epsilon$-closed, therefore
$\N.\epsdst~\epsilon\texttt{trans} \in \X$ because
$\N.\epssrc~\epsilon\texttt{trans}$ is in $X$. This concludes the construction of $\NtoD$.

@@

Next if the trace in $\D$ is defined via transition
$\X \overset{\c}{\to} (\D.\delta~\c~\X)$ then we may choose some state in
$\s \in \D.\delta~\c~\X$. Recursively, $\DtoN$ sends the trace rooted at
$\D.\delta~\c~\X$ to one rooted at $\s$. By definition of $\D.\delta~\c~\X$,
there exists some $\s' \in \X$ such that $\s'$ transitions to $\s$ by some number of
$\epsilon$-transitions followed by a transition labeled by $\c$.
We order all
such traces of this form and choose the smallest one, which is then linearly
combined with the parse of $\c$ to build an $\N$ trace. This concludes the
definition of $\DtoN$, thus showing that $\Trace_{\N}~\N.\init~\true$ and $\Trace_{\D}~\D.\init~\true$ are
weakly equivalent.
\end{proof}

@@

$\Dyck$ is a grammar over the alphabet $\{ \stringquote{(}, \stringquote{)} \}$
and is defined in \cref{fig:dyckinductive}. The $\nil$ constructor shows that
the empty string is balanced, and the $\balanced$ constructor builds a balanced
parse by wrapping an already balanced parse in an additional set of parentheses
then following it with another balanced parse.

@@

\subsection{Turing Machines}
\newcommand{\Turing}{\mathtt{Turing}}
\newcommand{\ttt}{\mathtt{t}}
\newcommand{\T}{\mathtt{T}}
\newcommand{\acc}{\mathtt{acc}}
\newcommand{\rej}{\mathtt{rej}}
\newcommand{\Tape}{\mathtt{Tape}}
\newcommand{\parseTape}{\mathtt{parseTape}}
\newcommand{\snoc}{\mathtt{snoc}}
\steven{This subsection is still a WIP}
%% 5. Expressiveness: Turing Machines
The expressive power of \theoryabbv spans the entire Chomsky hierarchy, as we may
encode the accepting traces of Turing machines as linear types.

@@

where $\mathsf{TuringTrace}$ is a nonlinear type denoting terminating traces
through the Turing machine. Note that $\mathsf{run}$ returns a $\mathsf{Maybe}$
type in a fuel parameter. Turing machines may not terminate. If they do
terminate within $\mathsf{fuel}$-many steps, then we return the trace along with
the current tape and a boolean denoting if the trace is accepting or rejecting.
If the trace does not terminate within the allotted amount of fuel, then we
timeout and return nothing.

@@

Parse transformers have an obvious notion of composition and identity
transformer. This defines a \emph{category}.
\begin{definition}
  Define $\Grammar$ to be the category whose objects are formal
  grammars and morphisms are parse transformers.
\end{definition}
This category is equivalent to the slice category $\Set/\Sigma^*$ and
as such is very well-behaved. It is complete, co-complete, Cartesian
closed and carries a monoidal biclosed structure. We will use these
structures to model the linear types, terms and equalities in
\theoryabbv.
%% \begin{theorem}
%%   The category $\Grammar$ is complete, co-complete and cartesian
%%   closed. Furthermore, it is monoidal biclosed with the monoidal
%%   product given by
%%   \[ (A \otimes B) w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A w_1 \wedge b \in B w_2\} \]
%%   and unit by
%%   \[ I w = \{ () \pipe w = \varepsilon \} \]
%% \end{theorem}

@@

For reasons of brevity, we present only a selection of cases in
\Cref{fig:semantics}. The full details are in the appendix.\max{TODO:
  put the full details in the appendix}.
\max{TODO: how to describe $\mu$ succinctly?}\max{make sure $\lto$ and $\tol$ are right}
\begin{figure}
  \begin{align*}
    \sem{c}\gamma\,w &= \{ c | w = c\}\\
    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
    \sem{\SPF\,X}\gamma &= \{ F : \textrm{Functor}(\textrm{Grammar}^{\sem{X}\gamma}, \textrm{Grammar}) \pipe F \textrm{ preserves $\omega$-colimits.} \}\\
    \sem{\el(F)}\gamma\,G &= \sem{A}\gamma\,G\\
    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:adamek}}
  \end{align*}
  \caption{Grammar Semantics}
\end{figure}

@@

\max{TODO: spelling}
\begin{theorem}[Adamek]
  If a functor $F : \mathcal C \to \mathcal C$ preserves
  $\omega$-colimits and $\mathcal C$ has an initial object and
  $\omega$-colimits, then $F$ has an initial algebra given by the
  colimit of the chain
  \[ 0 \to F(0) \to F(F(0)) \to \cdots \]
\end{theorem}

@@

\steven{Should discussion of this connection to Lambek calculus be spread
  throughout the paper?}

Lambek later explicitly connects his syntactic calculus to the structure of a
non-commutative biclosed monoidal category \cite{lambek1988categorial}. There is
a rich line of subsequent research after these revelations by Lambek, including
the study of pregroup grammars
\cite{coeckeMathematicalFoundationsCompositional2010}, the presentation of
abstract categorial grammars \cite{degrooteAbstractCategorialParsing2015}, and
even Lambek calculus with dependent types \cite{luoLambekCalculusDependent}. As
far as the authors can tell, these works seem to be primarily interested in natural
language semantics in the style of Montague grammars, rather than parsing formal
languages. That is, they seem to try to give a functorial semantic
interpretation of natural language.

@@

Also mention here that \cite{luoLambekCalculusDependent} seems similar at first
glance, but isn't as similar as it might look because the judgments and
dependent types seem to
take on a different interpretation of ours. Say something here like, ``as we
understand it, it differs in x,y,z ways''

\steven{TODO Edit the above}
\steven{Even though the above sounds bad, are there Lambek/categorial things that are missed?}

@@

\steven{TODO add Lambek calculus/categorial grammars here}

@@

    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
% %
%     \and
%
%     \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
% %
%     \and
% %
%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} {e \otimes e'} e'' \equiv e'' \{ e/a , e'/b \} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} e c : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(M, e)} {e'} \equiv e' \{ M/x , e/a \} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
    \and
    %
    \inferrule
    {~}
    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerin {e}} \equiv e : A}
    %
    \and
    %
    \inferrule
    {~}
    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e : A}
\end{mathpar}
  \caption{Judgmental equality for linear terms}
  \label{fig:jdgeq}
\end{figure}
\steven{Do we need inverse for $\ltonl A$ in judgmental equality figure?}
\steven{Do we need to say anything about the equalizer proofs being unique in
  judgmental equality figure?}
